---
title: "minfi methylation pipeline"
author: "Mikhail Dozmorov"
date: "Spring 2018"
output:
  beamer_presentation:
    # colortheme: seahorse
    colortheme: dolphin
    fig_caption: no
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    # theme: boxes
    theme: AnnArbor
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## `minfi` - Analyze Illumina Infinium DNA methylation arrays

- Reads Illumina’s 450k array raw data (IDAT files) into R
- Performs QC and normalization
- Identifies differential methylation positions (DMP)

```{r eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("minfi")
biocLite("minfiData")
```
```{r message=FALSE, warning=FALSE}
library(minfi)
library(minfiData)
```

\tiny https://bioconductor.org/packages/release/bioc/html/minfi.html

## Methylation data

```{r}
baseDir <- system.file("extdata", package = "minfiData")
list.files(baseDir)
targets <- read.metharray.sheet(baseDir)
RGset <- read.metharray.exp(targets = targets)
pd <- pData(RGset) ## phenotypic data
```

## QC: Beta values are expected to cluster around 0/1.

```{r message=FALSE, fig.height=4}
densityPlot(RGset, sampGroups = pd$Sample_Group, main = "Beta", xlab = "Beta")
```

## QC: Beta values are expected to cluster around 0/1.

```{r fig.height=4}
par(oma=c(2,10,1,1))
densityBeanPlot(RGset, sampGroups = pd$Sample_Group, sampNames = pd$Sample_Name)
```

## Normalization

Different methods for normalization have been proposed and still being developed

- Dye-bias adjustment
- Probe type I and II adjustment

```{r}
MSet.norm <- preprocessIllumina(RGset, bg.correct = TRUE, 
                                normalize = "controls", 
                                reference = 2)
```

&nbsp;

\tiny Yousefi P. et. al. "**Considerations for normalization of DNA methylation data by Illumina 450K BeadChip assay in population studies**" Epigenetics 2013 http://www.tandfonline.com/doi/abs/10.4161/epi.26037

## Multi-dimensional scaling (MDS) plot  {.smaller}

```{r fig.height=3.5}
mdsPlot(MSet.norm, numPositions = 1000, sampGroups = pd$Sample_Group, sampNames =pd$Sample_Name)
```

Similar to PCA, useful to identify outlier samples.

## Getting M-values

```{r}
# A small subset to speed up the demo:
mset <- MSet.norm[1:20000,]
# Getting the M values:
M <- getM(mset, type = "beta", betaThreshold = 0.001)
```

## QC: M values should show the level of methylation centered around 0

Beta values $\le 0.001$, or $\ge 0.999$ are truncated to avoid numerical issues.

```{r fig.height=3.5}
# Look at the density distribution
par(oma=c(2,10,1,1))
densityPlot(M, sampGroups = pd$Sample_Group, sampNames = pd$Sample_Name)
```

## Differentially methylated positions

```{r}
dmp <- dmpFinder(M, pheno=pd$Sample_Group, type="categorical")
head(dmp)
```

Rows ordered by p-value.

## Plotting methylation levels

```{r fig.height=4}
cpgs <- rownames(dmp)[1:4]
par(mfrow=c(2,2))
plotCpg(mset, cpg=cpgs, pheno=pd$Sample_Group)
```



